### 2013-08-19 doc/dissertation/document/chapters/introduction/introduction.tex
starting to incorporate some of Pauls initial comments

\chapter{Introduction}
\label{ch:introduction}
In nearly all high-fidelity nuclear reactor simulations, linear and
nonlinear transport problems are a primary focus of study. Neutronics,
fluid flow, and heat transfer are among the predominate physics that
fall into this category and simulation technologies to solve these
problems make up much of the analysis suite used for modern
engineering calculations. Recent focus on multiple physics systems in
the nuclear reactor modeling and simulation community adds a new level
of complexity to common linear and nonlinear systems as solution
strategies change when they are coupled to other problems
\cite{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these transport systems as a means to
capture effects not modeled by coarser methods.

@@

\subsection{Solutions for the $SP_N$ Equations}
\label{subsec:spn_motiviation}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
reactor neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

@@

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment in many forms with both
variational and asymptotic derivations performed as a means of
verification \cite{olbrant_asymptotic_2013}. In addition, these
equations have been more rigorously studied as solution methods to MOX
fuel problems and have been shown to provide accurate solutions
\cite{brantley_simplified_2000}. With this mathematical literature to
provide a solid numerical footing for the method, we look at its
application to today's challenge problems in neutron transport for
fission reactor analysis. The reduction in numerical complexity of
current deterministic solution methods using the $S_N$ approximation
could mean significant savings in both compute time and memory
required. In addition, the characteristics of the solution to the
transport problem for a steady state reactor core permit diffusion
theory to be used; a staple of the nuclear industry since its
inception. Therefore, if diffusion theory is applicable, then finer
grained solutions that capture more of the physics contained in the
transport equation should be possible with the $SP_N$ method. In doing
so, we also expect from the literature to obtain computed responses on
the order of accuracy we would expect from an appropriately
discretized $S_N$ method at a fraction of the cost.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems \cite{evans_simplified_2013}. By fully forming
these equations and formulating them as a linear algebra problem
instead of using the explicit iterative methods of the past, we now
have access to all of the modern advancements in computational linear
algebra including Krylov solvers for asymmetric systems and algebraic
preconditioning methods.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems are tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis of the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow, then we are motivated to advance the
solution of complex physics problems exploiting leadership class
levels of parallelism.

>>>>

\chapter{Introduction}
\label{ch:introduction}
In nearly all high-fidelity nuclear reactor simulations, linear and
nonlinear transport problems are a primary focus of study. Neutronics,
fluid flow, and heat transfer are among the predominant physics that
fall into this category and simulation technologies to solve these
problems make up much of the analysis suite used for modern
engineering calculations. Recent focus on multiple physics systems in
the nuclear reactor modeling and simulation community adds a new level
of complexity to common linear and nonlinear systems as solution
strategies change when they are coupled to other problems
\cite{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these transport systems as a means to
capture effects not modeled by coarser methods.

@@

\subsection{Solutions for the $SP_N$ Equations}
\label{subsec:spn_motiviation}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, contain large numbers of degrees of freedom, or both,
requiring tremendous computational resources to generate an adequate
solution. Modern high fidelity deterministic methods for large scale
problems are commonly variants on the discrete ordinates ($S_N$)
method \cite{evans_denovo:_2010}. For fission reactor neutronics
simulations, the $S_N$ method requires potentially trillions of
unknown angular flux moments to be computed to achieve good accuracy
for the responses of interest
\cite{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

@@

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment in many forms with both
variational and asymptotic derivations performed as a means of
verification \cite{olbrant_asymptotic_2013}. In addition, these
equations have been more rigorously studied as solution methods to MOX
fuel problems and have been shown to provide accurate solutions
\cite{brantley_simplified_2000}. With this mathematical literature to
provide a solid numerical footing for the method, we look at its
application to today's challenge problems in neutron transport for
fission reactor analysis. The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems \cite{evans_simplified_2013}. By fully forming
these equations and formulating them as a linear algebra problem
instead of using the explicit iterative methods of the past, we now
have access to all of the modern advancements in computational linear
algebra including the use of Krylov solvers for asymmetric systems
without the need to implement the action of the physics operator on a
vector and easy application of algebraic preconditioning methods for
improved convergence properties and time to solution.

@@

The computational resources required to solve the neutronics and fluid
dynamics problems can be tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example, was performed as part of analysis of the
Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\cite{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \cite{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance solution techniques
in neutronics and fluid flow, then we are motivated to advance the
solution of complex physics problems exploiting leadership class
levels of parallelism.

### 2013-08-19 doc/dissertation/document/chapters/mc_background/mc_background.tex
pushing full draft of disseration
fixing typo in adjoint constraints
recovering from failed rebase
starting to incorporate some of Pauls initial comments

\chapter{Monte Carlo Synthetic Acceleration Methods}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's.

@@

%%---------------------------------------------------------------------------%%
\section{Neumann-Ulam Method}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods by seeking a solution
to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
we can use diagonal splitting in a similar manner as the stationary
method in Eq~(\ref{eq:linear_split_equation2}) to define the following
operator\footnote{It should be noted that non-diagonal splittings have
  been recently explored \cite{srinivasan_monte_2010} and have the
  potential to improve efficiency. However, it was observed in this
  work that this type of splitting did not improve performance in the
  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
  problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

In cases where we seek only approximate solutions, we need only to
perform a predetermined number of random walks in order to generate an
approximation for $\ve{x}$. If we are only to approximate the
solution, we also need conditions by which we may terminate a random
walk as the Neumann Ulam decomposition defined by
Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
that approaches, but never reaches zero. We do this by noticing that
the factors added to Eq~(\ref{eq:direct_permutation_weight}) will
become diminishingly small due to their definition in
Eq~(\ref{eq:direct_weight}) and therefore their contributions to the
solution estimate will become negligible. Using this, we choose
terminate a random walk sequence with a \textit{weight cutoff}, $W_c$,
that is enforced when $W_m < W_c$ for a particular random walk
permutation.

@@

and applying Eq~(\ref{eq:direct_permutation_contribution}):

@@

  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k W_{m}^2 b_{i_m}^2
  + 2 \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} \Big)-

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with a single history at each state in the domain, the
high variance prevents a precise solution from being obtained although
we begin to see the solution take shape as expected. At 1000 histories
per state, enough tallies have been made to generate a reasonable
estimate for the structure of the solution. It is interesting to note
here that as the statistical uncertainty is reduced at each grid point
in the domain, the solution is resolved in a certain sense, analogous
to the convergence of a traditional iterative method.
\clearpage

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}):
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
  \ve{y} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

\begin{figure}[p!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. artificial
      absorption probability for a $200 \times 200$ square mesh and
      40,000 histories.}  \textit{The addition of absorption is
      detrimental to iterative performance due to both the shortening
      of the random walks as well as the modification of the
      transition weight.}}
  \label{fig:absorption_iters}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
      probability for a $200 \times 200$ square mesh and 40,000
      histories.}  \textit{The reduced random walk length speeds up
      the calculation with a fixed number of histories. Too much
      artificial absorption increases iterations rapidly giving an
      increasing CPU time.}}
  \label{fig:absorption_time}
\end{figure}
\clearpage

>>>>

\chapter{Monte Carlo Synthetic Acceleration\\ Methods}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's.

@@

%%---------------------------------------------------------------------------%%
\section{Neumann-Ulam Method}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods by seeking a solution
to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
we can use diagonal splitting similar to the stationary method in
Eq~(\ref{eq:linear_split_equation2}) to define the following
operator\footnote{It should be noted that non-diagonal splittings have
  been recently explored \cite{srinivasan_monte_2010} and have the
  potential to improve efficiency. However, it was observed in this
  work that this type of splitting did not improve performance in the
  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
  problems.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this Neumann series to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation by truncation will therefore lead
to an approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering the $i^{th}$
component of the solution vector:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
sequence permutation. We can generate these sequences of transitions
through Monte Carlo random walks by assigning them both a probability
and weight. As a reinterpretation of the iteration matrix, we then
form the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence and therefore truncate the Neumann series it is
approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
use the provided linear operator $\ve{A}$ to form the Neumann Ulam
decomposition while the adjoint method will use the adjoint linear
operator ($\ve{A}^T$ for real-valued systems) to form the
decomposition.

@@

In cases where we seek only approximate solutions, we need only to
perform a predetermined number of random walks in order to generate an
approximation for $\ve{x}$. If we are only to approximate the
solution, we also need conditions by which we may terminate a random
walk as the Neumann Ulam decomposition defined by
Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
that approaches, but never reaches zero. We do this by noticing that
if the problem is convergent, the factors added to
Eq~(\ref{eq:direct_permutation_weight}) will make $W_m$ become
diminishingly small due to their definition in
Eq~(\ref{eq:direct_weight}) and therefore the contribution of the
history to the solution estimate will become negligible. Using this,
we choose terminate a random walk sequence with a \textit{weight
  cutoff}, $W_c$, that is enforced when $W_m < W_c$ for a particular
random walk permutation.

@@

and applying Eq~(\ref{eq:direct_permutation_contribution}) and
expanding the square of the summation:

@@

  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k \Big[ W_{m}^2 b_{i_m}^2
  + 2 \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} \Big] \Big)-

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with a single history at each state in the domain, the
high variance prevents a precise solution from being obtained although
we begin to see the solution take shape as expected. It is interesting
to note here that as the statistical uncertainty is reduced at each
grid point in the domain, the solution is resolved in a certain sense,
analogous to the convergence of a traditional iterative method.
\clearpage

%%---------------------------------------------------------------------------%%
\section{Adjoint Monte Carlo Method}
\label{sec:adjoint_mc}
Often a more useful form, an alternative to forward Monte Carlo matrix
inversion is the adjoint method. We begin by defining the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
mirror Eq~(\ref{eq:richardson_split}):
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
  \label{eq:adjoint_split_system}
\end{equation}
As was required for convergence with the direct method using
Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
therefore has the same spectral radius. By defining the following
inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
Using these definitions, we can derive an estimator from the adjoint
method that will also give the solution vector, $\ve{x}$. As with the
direct method, we can acquire the adjoint solution by forming the
Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence, this time
forming the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. artificial
      absorption probability for a $200 \times 200$ square mesh and
      40,000 histories.}  \textit{The addition of absorption is
      detrimental to iterative performance due to both the shortening
      of the random walks as well as the modification of the
      transition weight.}}
  \label{fig:absorption_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
      probability for a $200 \times 200$ square mesh and 40,000
      histories.}  \textit{The reduced random walk length speeds up
      the calculation with a fixed number of histories. Too much
      artificial absorption increases iterations rapidly giving an
      increasing CPU time.}}
  \label{fig:absorption_time}
\end{figure}
\clearpage
